# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
#load data
activityData <- read.csv('activity.csv')
#remove rows with NA values for steps column
activiyDataNoNA <- na.omit(activityData)
```


## What is mean total number of steps taken per day?

Below is a histogram that plots the total number of steps taken each day.
```{r}
library(plyr)
stepsPerDay <- ddply(activiyDataNoNA, .(date), summarise, totalSteps=sum(steps))
hist(stepsPerDay$totalSteps)
```

The mean of the total number of steps taken per day is `r mean(stepsPerDay$totalSteps)`.  

The median of the total number of steps taken per day is `r median(stepsPerDay$totalSteps)`.  

## What is the average daily activity pattern?

Below is a time series plot of the average number of steps per 5-minute time interval. The averages are calculated accross all days.

```{r}
avgStepsPerInterval <- ddply(activiyDataNoNA, .(interval), summarise, avgSteps=mean(steps))
plot(avgStepsPerInterval$interval, avgStepsPerInterval$avgSteps, type="l")
```

Now let's find the interval on average with the most steps.

```{r results='hide'}
intervalWithMaxSteps <- avgStepsPerInterval[avgStepsPerInterval$avgSteps == max(avgStepsPerInterval$avgSteps),1]
```

On average the interval with the most steps is the interval `r intervalWithMaxSteps`.

## Imputing missing values

We are going to estimate how the missing values in the original data affect the results we get back.  

First we will calculate the total number of missing values in the original dataset.
```{r results='hide'}
numNAValues <- nrow(activityData[is.na(activityData$steps),])
```

So the total number of rows with missing steps values is `r numNAValues`.

Next, we will fill in estimates for the missing steps values. We are going to fill in any missing steps value with the mean value over all days for that interval.

```{r results='hide', cache=TRUE}
newActivityData <- activityData
for (i in 1:nrow(newActivityData)) {
	row <- newActivityData[i,]
	
	if (is.na(row$steps)) {
		#steps is an NA value so get the avg steps for time interval
		newActivityData[i,"steps"] <- avgStepsPerInterval[
			avgStepsPerInterval$interval == row$interval,"avgSteps"]
	}
}
```

Now we will plot another histogram of the number of steps taken each day but using this new data with missing values filled in:

```{r}
stepsPerDay <- ddply(newActivityData, .(date), summarise, totalSteps=sum(steps))
hist(stepsPerDay$totalSteps)
```

The new mean of the total number of steps taken per day is `r mean(stepsPerDay$totalSteps)`.  

And the new median of the total number of steps taken per day is `r median(stepsPerDay$totalSteps)`.  

As you can see the new mean after imputing missing values is the same as the previous mean calculated when we removed all NA values.  

The new median however differs slightly and is now the same as the mean where it was slightly smaller before.



## Are there differences in activity patterns between weekdays and weekends?
